---
title: "Get Started"
author: "Paradigm4"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette

---

```{r, child='shared/_setup.Rmd'}
```


# Connect to SciDB

```{r, eval=FALSE}
library(arrayop) # optional step to leave out the `arrayop::` prefix.
# First, load user name and token from an authentication file, which is also compatible with iquery.
auth = yaml::yaml.load_file("~/.scidb_auth")
# Then connect to scidb. Params host, username and token are all single strings and can be provided in any way
conn = db_connect(host = "localhost", username = auth[["user-name"]], token = auth[["user-password"]])
```

Once we have connected to scidb with `db_connect`, we can get save the connection instance as shown above.
Or call `arrayop::get_default_connection()` anywhere in code, which always return the db connection.

```{r}
identical(get_default_connection(), conn) # the exact same thing!
print(conn) # print out `username@host [scidb version]` if the connection is succesful
```


# Get array_op instances

We use the term `array_op` to denote two types of arrays:

  1. persistent array in SciDB.
  2. transient array, ie. any numbers of potentially nested operations on one or multiple arrays. 
  E.g. filter on one array; or join two arrays then filter the joined.
  
Semantically, both arrays are the same, and support the same [set of verbs](`ArrayOpBase`). 
  

```{r}
# Upload a data frame to scidb
arr_uploaded = conn$upload_df(CO2, name = "test_CO2") # datasets::CO2

# Or just create a new array
arr_created = conn$create_array(
  name = "test_CO2_empty",  
  schema_template = "<Plant:string,Type:string,Treatment:string,conc:double,uptake:double> [i]"
)

# Retrieve an existing array (the one we just uploaded)
arr_existing = conn$array("test_CO2")

# Compile data.frame into a transient array
arr_compiled = conn$compile_df(head(CO2,4))

# Create a transient array by AFL
arr_from_afl = conn$array_from_afl(arr_existing$filter(conc > 800)$to_afl())

# Store an AFL statement into a persistent array
arr_stored = conn$array_from_stored_afl(
  # afl_str = "filter(test_CO2, conc > 800)" # equivalent to code below
  afl_str = arr_existing$filter(conc > 800)$to_afl(), 
  save_array_name = "test_CO2_stored", .gc = F # prevent from being destroyed by R's garbage collected
)
```

Let's inspect these arrays!

```{r}
arr_uploaded$to_afl()
arr_uploaded$to_schema_str()
arr_uploaded$is_persistent()
arr_uploaded$array_meta_data()
arr_uploaded$summarize_array()
arr_uploaded$row_count()

str(arr_created)
str(arr_existing) # same as arr_uploaded

arr_compiled$to_afl()
arr_compiled$to_schema_str()
arr_compiled$is_persistent()

arr_from_afl$to_afl()
arr_from_afl$row_count()

str(arr_stored)
arr_stored$row_count()
```

# Download array_op as data.frame

Call `to_df` method on any `array_op` instance to get a data frame with all its attributes; 
or `to_df_all` to get all fields (both dimensions and attributes).

```{r}
arr_stored$to_df()
arr_stored$to_df_all() # Notice the dimension column [i]
```


**Common array operations**

We can create array_ops on the fly and download them as data frames.
Each array operation also returns an array_op instance. 
So we can chain up the operations in a `dplyr` like data flow. 
All the array operation verbs are named following `dplyr` convention where appropriate,
e.g. `filter`, `inner_join`, `mutate`, `group_by` and `summarize` etc.


# Filter

```{r}
# Comma separated expressions are joined with logical AND
arr_existing$filter(conc > 80, uptake > 40)$to_df()
# Equivalent to above
# arr_existing$filter(conc > 80 & uptake > 40)$to_df()

arr_existing$filter(conc <= 100 | conc > 900)$to_df() # logical OR

# Filter on string fields
# %like% takes regex as the right operand; all other verbs treat strings literally.
arr_existing$filter(Plant %contains% "3", Plant %starts_with% "Q", Treatment %contains% "chill", Treatment %like% "chill[a-z]{2}$")$to_df()
```

# Mutate/Transmute

## By expressions

As shown below, field names can be either quoted or not quoted. 
But the field expression must be a single string. 
Unlike R, SciDB uses single quotes for string literals. E.g. `'at'` is single-quoted.

```{r}
# Modify existing fields
result = arr_stored$mutate("Type" = " 'at' + Type ", conc = "conc/1000")
result$to_afl()
result$to_df()

# Add new fields
result = arr_stored$mutate("Plant_Type" = "Plant + '-' + Type", "feel" = "iif(Treatment = 'chilled', 'cold', 'ok')")
result$to_afl()
result$to_df()
```

`mutate` operation keeps the original fields unless we modify them. 

`transmute` allows us to explicitly specify which fields to keep.

```{r}
result = arr_stored$transmute("Plant", "Plant_Type" = "Plant + '-' + Type", "feel" = "iif(Treatment = 'chilled', 'cold', 'ok')")
result$to_afl()
result$to_df()
```

## By arrays

`mutate` and `transmute` are useful if the mutated fields are in expressions. 
Another common use case is mutating fields of an existing array by another array, rather than by expressions.

```{r}
dataArray = conn$array_from_df(data.frame(Plant = c("Qn1", "Qn2", "Qn3"), Treatment = "treated-mutated"))
result = arr_stored$mutate_by(dataArray, keys = "Plant", updated_fields = "Treatment")
result$to_afl()
result$to_df()
```

# Array Mutability 

**NOTE**: the source arrays remain the same in examples above. All mutate operations
just created **mutated data copies** of the source.

To modify array content, we can `update` or `overwrite` the source array with **mutated data copies**.

```{r}
dataArray = conn$array_from_df(data.frame(Plant = c("Qn1", "Qn2", "Qn3"), Treatment = "treated-mutated"))
mutated = arr_stored$mutate_by(dataArray, keys = "Plant", updated_fields = "Treatment")

# Persist data changes to an existing array
mutated$update(arr_stored)$execute()
arr_stored$to_df() # Only matched data cells are updated

mutated$overwrite(arr_stored)$execute()
arr_stored$to_df() # Caution: Overwrite all array content!!!
```

# Join

Three joins are supported: `inner_join`, `left_join` and `right_join`.

First, let's populate two arrays.

```{r}
leftDf = data.frame(lda = 1:5, db = 11:15, fa = LETTERS[1:5],  lfb = 3.14 * 1:5)
rightDf = data.frame(rda = 3:10, db = 13:20, fa = LETTERS[3:10], rfb = 3.14 * 3:10)

leftDf
rightDf

# At least one input to 'cross_join' must have a specified chunk size.
# So we need to persist the arrays
L = conn$array_from_df(leftDf, "<fa:string, lfb:double> [lda;db] ", force_template_schema = T)$persist(.gc = F)
R = conn$array_from_df(rightDf, "<fa:string, rfb:double> [rda;db] ", force_template_schema = T)$persist(.gc = F)
```

Then, start with a vinilla join where join keys are inferred from array fields.
In this case, `fa` and `db` are join keys since they exist in both arrays.

```{r}
L$inner_join(R)$to_df()
L$left_join(R)$to_df()
L$right_join(R)$to_df()
```

We can specify which fields to keep with `array_op$select`. 
By default, all fields are selected, except that the join keys only exist once in the result even if they are selected on both operands.

```{r}
L$select("lfb", "lda")$
  inner_join(R)$to_df()

L$inner_join(R$select("rda", "rfb"))$to_df()

L$select("lfb", "db")$inner_join(R$select("rfb"))$to_df()
```

We can specify the join keys and aliases explicitly.
Array aliases are important to disambiguate overlapping fields.

```{r}
# `db` as join key, `fa` exists in both arrays and will be suffixed with array aliaes in the result
L$left_join(R, on_both = c('db'), left_alias = "_LL", right_alias = "_RR")$to_df()

# different join keys
L$right_join(R, on_left = "lda", on_right = "rda", left_alias = "_LL", right_alias = "_RR")$to_df()
```

## Join mode

By default, join is implemented by SciDB `equi_join` operator, which works great on either array dimensions and/or attributes. 
But `equi_join` has one drawback as it materializes results. This can be underiable in a complex nested join.

We can choose a different join mode, `cross_join`, to avoid result materialization. 

Two constraints with `cross_join` mode:

  1. join keys must be dimensions on both operands.
  1. only `inner_join` is supported.

```{r}
L$inner_join(R, 
             on_left = c('db', 'lda'), 
             on_right = c('db', 'rda'), 
             join_mode = 'cross_join', left_alias = "_LL", right_alias = "_RR")$
  to_df()
```

## Custom settings

```{r}
result = L$inner_join(R, 
             settings = list(
               algorithm = "'hash_replicate_right'", # string setting item
               hash_join_threshold = 1024 # numeric setting item
         ))
result$to_afl()
result$to_df()
```



# Aggregation
```{r}
# Tip: Call `dbutils$db_aggregates()$to_df()` to see supported aggregate functions
arr_existing$
  group_by("Plant", "Type")$
  summarize("count" = "count(*)",
            "uptake_sd" = "stdev(uptake)",
            "avg(conc)", # unamed fields are named by SciDB
            "max(uptake)")$
  to_df()

# Aggregation result is still an array_op. So we can carry on.
arr_existing$
  group_by("Plant", "Type")$
  summarize("count" = "count(*)",
            "uptake_sd" = "stdev(uptake)",
            "avg(conc)", 
            "max(uptake)")$
  filter(uptake_sd > 8)$ # requires uptake standard deviation > 8
  to_df()
```

# Versions

Since we just modified `arr_stored` several times. 
Let's check out how many versions it has so far and delete the version history.

```{r}
arr_stored$versions()
arr_stored$remove_versions()
arr_stored$versions()
```





# Raw AFL

Hopefully `arrayop` captures all your common SciDB use cases.
In case something is missing, we can run raw AFL statements directly on a ScidbConnection object.

To run AFL statements (in string aka. length-1 R character vector format), use one of the methods of a `ScidbConnection`:

  1. `query` if we expect results (array attributes as R data frames).
  1. `query_all` if we need both array dimensions and attributes.
  1. `execute` if we do not expect results 
  
```{r}
conn$query("list('operators')")
conn$query("op_scidbversion()")
```

Let's create an array `temp_array` in the `public` namespace for later examples.
```{r}
conn$execute("create array public.temp_array <a:int32, b:string, c:bool> [z]")
conn$query("show(temp_array)") # query(...) returns result attribtues only
conn$query_all("show(temp_array)") # query_all(...) returns result dimesions and attributes
```

## Chain AFL operators with pipes

We can pipe multiple SciDB operators in `arrayop::afl` function which returns an AFL string.

Depending on the number of operands:

  - `afl(a | b)` is converted to `b(a)`
  - `afl(a | b(c))` is converted to `b(a, c)`

```{r}
conn$query(afl(
  "'operators'" | list | filter("library != 'scidb'") | limit(10)
))
```


# Remove arrays

**NOTE:** removed arrays cannot be recovered. 
Always double check and back up if necessary before entering the danger zone.

```{r}
conn$execute("remove(temp_array)") # manually remove

for(x in c(arr_uploaded, 
           arr_created, 
           arr_existing,
           arr_compiled, 
           arr_stored, 
           L, R
           )){
  if(x$exists_persistent_array()) { # can only remove persistent arrays
    x$remove_self()
  }
}

```

