---
title: "Semi Join"
author: "Paradigm4"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette

---

```{r, child='shared/_setup.Rmd'}
```


# Create an array

We first create a `RefArray` with content from `ArrayContent` data frame
```{r}
ArrayContent = data.frame(
  da=1:20, db=101:120,
  lower = letters[1:20], 
  upper = LETTERS[1:20],
  f_int32 = -20:-1, 
  f_int64 = 1:20 * 10.0, 
  f_bool = c(T,NA,F,NA,F), 
  f_double = c(3.14, 2.0, NA, 0, -99)
)

RefArray = conn$
  array_from_df(ArrayContent, 
  template =  "<lower:string COMPRESSION 'zlib', upper:string, f_int32:int32, f_int64:int64, f_bool: bool, f_double: double> 
      [da=0:*:0:*; db=0:*:0:*]", 
  force_template_schema = T)$ # Ensure RefArray has the exact schema as the template
  persist(.gc = FALSE)

RefArray$to_afl()
RefArray$to_df_all()
```

# Overview

We can get an arrayOp with the same schema as `RefArray` but filtered content.
The filtering criteria can be from either an R data frame or an arrayOp instance.

Let's define a function to show the semi join result arrayOp afl and download 
the data frame.

```{r}
show = function(result_array_op){
  print(result_array_op$to_afl())
  result_array_op$to_df_all() # also download dimensions
}
```

# Filter mode

When the filtering criteria is from an R data frame and its cell count is relative
small to fit in a scidb build literal, we can use `mode = 'filter'`.

## By one field
```{r}
RefArray$semi_join(
  data.frame(da = c(3,5,8,11)), # number of cells <= filter_threshold
  filter_threshold=10, upload_threshold=20) %>%
  show

# No matched cells
RefArray$semi_join(
  data.frame(da =  c(-10, -11, -12)),
  filter_threshold=10, upload_threshold=20) %>%
  show
```
## By two fields
```{r}
RefArray$semi_join(
  data.frame(
    f_int32 =  c(-20, -17, -16),
    lower = c("no_match", "d", "e")
  ),
  filter_threshold = 10,
  upload_threshold = 20
) %>%
  show


RefArray$semi_join(
  data.frame(da =  c(-1, 4, 5),
             lower = c("no_match", "d", "e")),
  filter_threshold = 10,
  upload_threshold = 20
) %>%
  show

# Set lower/upper bounds to range query array dimensions
RefArray$semi_join(
  data.frame(da_low = c(1, 3), da_hi = c(5, 8)),
  lower_bound = list('da' = 'da_low'),
  upper_bound = list('da' = 'da_hi'),
  filter_threshold = 10,
  upload_threshold = 20
) %>%
  show

# Set lower/upper bounds to on different dimensions
RefArray$semi_join(
  data.frame(da = c(1, 3), db = c(105, 108)),
  lower_bound = list('da' = 'da'),
  upper_bound = list('db' = 'db'),
  filter_threshold = 10,
  upload_threshold = 20
) %>%
  show
```
# cross_between mode

When the query data frame has cell count greater than `filter_threshold`, but 
smaller than `upload_threshold`, the data frame is converted to a build literal.
Or if its greater than `upload_threshold`, the data frame is uploaded as a persistent 
scidb array first.

In both case, if the arrayOp from the query data frame has more than one attribute,
a `cross_between` mode is automatically chosen.

```{r}
RefArray$semi_join(
  data.frame(da = c(1:5, -1), db = c(101:105, -1)),
  filter_threshold = 10,
  upload_threshold = 20
) %>%
  show

RefArray$semi_join(
  data.frame(da_low = c(1,3), da_hi = c(5, 8)),
  lower_bound = list(da = 'da_low'), 
  upper_bound = list(da = 'da_hi'),
  filter_threshold = 10,
  upload_threshold = 20
) %>%
  show

# The query is already an arrayOp, then the threshold is irrelevant
# cross_between mode is chosen if the query arrayOp has more than one 
# attribute
RefArray$semi_join(
  conn$array_from_df(
    data.frame(da_low = c(1, 3), da_hi = c(5, 8)),
    "<da_low:int64, da_hi:int64> [anything]"
  ),
  lower_bound = list(da = 'da_low'),
  upper_bound = list(da = 'da_hi'),
) %>%
  show


# Explicitly set `field_mapping` to an empty list if the matching fields are 
# supposed to be used as bounds instead of exact matching
RefArray$semi_join(
  data.frame(da = c(1, 3), db = c(105, 108)),
  lower_bound = list(da = 'da'),
  upper_bound = list(db = 'db'),
  field_mapping = list(),
  filter_threshold = 2,
  upload_threshold = 20
) %>%
  show

RefArray$semi_join(
  conn$array_from_df(
    data.frame(da = c(1, 3), db = c(105, 108)),
    template = RefArray
  ),
  lower_bound = list(da = 'da'),
  upper_bound = list(db = 'db'),
  field_mapping = list()
) %>%
  show

```

# index_lookup mode

If the query arrayOp has only one attribute, 'index_lookup' mode is automatically 
chosen unless otherwise provided.

```{r}
# by attribute
RefArray$semi_join(
  data.frame(lower = letters[5:15]),
  filter_threshold = 5,
  upload_threshold = 20
) %>% show

# by dimension
RefArray$semi_join(
  data.frame(da = 1:15),
  filter_threshold = 5,
  upload_threshold = 20
) %>% show

```

```{r, echo=FALSE}
# Clena up
RefArray$remove_array()
```

